Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-08-21 Chapter_0/04_Manuscript/Manuscript_Backup_2020-06-12.tex
first commit

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
\noindent \textbf{1. Background}: Protecting large portions of natural land
connecting natural reserves is an increasingly used strategy to maintain and
restore connectivity among wildlife subpopulations. One such initiative is the
Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA) in Africa, the
world's largest transboundary conservation area, spanning five countries and
520'000 km\textsuperscript{2}. While boundaries of such conservation areas are
often determined by expert opinion as well as socio-political needs and
constraints, the extent to which they match species' movement corridors is
rarely examined. This is mainly due to a lack of adequate empirical data,
particularly on movement behavior of wide-ranging species during dispersal.

\noindent \textbf{2. Methods}: Between 2011 and 2019, we collected
high-resolution GPS relocation data on 16 coalitions of dispersing African wild
dogs (\textit{Lycaon pictus}) from a free ranging population in northern
Botswana. We combined the collected data with relevant spatial covariates and
used integrated step selection analysis to estimate habitat preferences during
dispersal and create a permeability surface spanning the entire KAZA-TFCA. To
examine whether the KAZA-TFCA covers major movement corridors of this highly
mobile species, we compared landscape permeability across different regions of
our study area. We also calculated least-cost paths and corridors between
selected locations and verified that corridors are encompassed within the
boundaries of the KAZA-TFCA.

\noindent \textbf{3. Results}: We show that permeability within the KAZA-TFCA is
more than double compared to areas outside it. Furthermore, we report a
five-fold permeability difference among the five KAZA-TFCA countries. These
differences are mainly caused by different degrees of human activities across
regions, hampering dispersal movements. We further present a connectivity
network, consisting of least-cost paths and corridors. We thereby show that all
major movement corridors of wild dogs run within the KAZA-TFCA boundaries,
although some minor corridors remain outside formally protected areas.

\noindent \textbf{4. Synthesis and Applications}: Our work highlights the
importance of empirically assessing the adequacy of protected areas and the need
for international coordination that stem from different contributions to
connectivity across regions and countries. Moreover, the presented framework is
readily expandable to include additional species and regions in order to
highlight consistencies among potential dispersal corridors as well as dispersal
corridors outside our ecosystem.

\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. A sufficient degree of connectivity
enables dispersal, which in turn contributes to population viability through
gene-flow, the ability to reinforce and rescue small populations, to colonize or
recolonize unoccupied habitats, as well as to shift range in response to climate
change \citep{Brown.1977, Hanski.1998, MacArthur.2001, Frankham.2002,
Heller.2009, Leigh.2012}. Particularly in light of ongoing habitat fragmentation
and destruction worldwide \citep{Fahrig.2003, Lindenmayer.2013}, the protection
of natural or semi natural areas that serve as movement corridors and provide
connectivity has become an utmost task for conservation management
\citep{Heller.2009, Doerr.2011, Rudnick.2012, Cozzi.2020}. This has resulted in
an ever-growing number of large and often transboundary protected areas that are
set to facilitate dispersal between habitats \citep{Wolmer.2003}. Due to the
very nature of large scale protected areas, a thorough quantitative assessment
of their effectiveness and adequacy of their boundaries is challenging
\citep{Rudnick.2012}. In result, boundaries are typically drawn according to
expert opinion and socio-political needs rather than empirical information. In
comparison to qantitative assessments, however, such qualitative approaches have
revealed deficiencies in the past and potentially hold the risk of misdirecting
valuable and scarce conservation funds \citep{Pullinger.2010, Sawyer.2011}. An
accurate representation of movement corridors and dispersal routes based on
empirical data is thus fundamental to objectively assess how well boundaries of
existing or planned protected areas reflect natural movements of their protected
species \citep{Beier.2008, Doerr.2011, Rudnick.2012}.

@@

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts, but for
which systematically collected GPS relocation data on dispersing individuals are
unavailable. Once widespread across the entire Sub-Saharan continent, wild dogs
have been widely extirpated through human persecution, habitat destruction, and
disease outbreaks \citep{Woodroffe.2012}. As a result, the species has become
one of Africa's most endangered large carnivores, and currently only survives in
small, spatially scattered subpopulations \citep{Woodroffe.2012}. Within these
subpopulations, wild dogs form cooperative breeding packs of up to thirty
individuals \citep{Frame.1979, Fuller.1992, Creel.2002}, whose social structure
is strongly governed the process of dispersal \citep{McNutt.1996,
Woodroffe.2019, Behr.2020}. Both males and females disperse from their natal
pack, either alone or in same-sex dispersing coalitions, and search for
unrelated mates and a suitable territory to settle \citep{McNutt.1996,
Cozzi.2020, Behr.2020}. During dispersal, wild dogs can cover several hundred
kilometers, sometimes crossing national boundaries and areas that are strongly
influenced by humans \citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers during dispersal. The few studies that have
collected GPS relocation data during dispersal have shown that dispersers
quickly move over large distances \citep{Woodroffe.2019}, generally avoid
human-dominated landscapes \citep{Masenga.2016, Oneill.2020, Cozzi.2020} and
areas densely forested \citep{Oneill.2020}, but prefer proximity to water and
dirt roads \citep{Oneill.2020}.

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
used it to assess the adequacy of the KAZA-TFCA in securing connectivity for
this species. To achieve this, we estimated the degree of selection or avoidance
for environmental and anthropogenic landscape features during dispersal. We then
applied the obtained habitat preferences and predicted a permeability surface
spanning the entire KAZA-TFCA and we investigated how landscape permeability
varies regionally and internationally. Finally, we calculated least-cost paths
and corridors to identify major movement corridors, pinpoint dispersal hubs, and
verify that the KAZA-TFCA successfully covers these features.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E; elevation ca.
1'030 m; red dotted rectangle in \Cref{StudyArea}a) was outlined by a
rectangular bounding box stretching over 1.3 Mio km\textsuperscript{2} and
encompassing the entire KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA lies in the
basins of the Okavango and Zambezi rivers and includes parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia. With a total area of over 520'000
km\textsuperscript{2} it covers several already-existing national parks and
other protected areas. It constitutes the earth's largest transboundary
conservation area and is characterized by diverse landscapes, including savanna,
grassland, and dry or moist woodland habitats. Rainfall in the study area is
seasonal and lasts from November to March, whereas the rest of the year remains
dry \citep{Mendelsohn.2010}. The KAZA-TFCA also encloses the Okavango Delta,
which is one of the main strongholds of African wild dogs in southern Africa and
likely acts as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012, Cozzi.2013}. The Okavango Delta represents a highly
dynamic hydrological flood-pulsing system \citep{McNutt.1996, Wolski.2017}. The
extent of the flood in the delta greatly changes within and between years
depending on the amount of rain that descends from the catchment areas in Angola
and reaches the distal ends of the delta between July and August (see Figure S5
for an illustration of a typical flood pulse). The flood drastically affects
surrounding landscapes, so that during maximum extent (ca. 12'000
km\textsuperscript{2}) the delta becomes a patchy conglomerate of swamps, open
water, and islands, whereas these structures run dry when the flood retracts to
its minimum extent (ca. 5'000 km\textsuperscript{2}; \citealp{Gumbricht.2004,
Wolski.2017, McCarthy.1998}). Despite 36 NPs and other protected areas, there is
considerable human influence in some regions of the KAZA-TFCA, mainly
originating from farms, high human density, and road traffic.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' selection or avoidance for the above-mentioned spatial
covariates. In the iSSF framework, GPS relocations are coerced to  ``steps'',
where a step is defined as the connecting line between two consecutive GPS
relocations \citep{Turchin.1998}. Subsequently, habitat preferences are
estimated by contrasting covariates along realized steps to the same covariates
along alternative ``random''  steps \citep{Fortin.2005, Thurfjell.2014,
Avgar.2016}. Thus, we paired each realized step with 24 random steps. We
generated random steps by sampling turning angles from a uniform distribution
U(\(-\pi, +\pi\)) and step lengths from a gamma distribution that was fitted
using realized steps \citep{Avgar.2016}. Together, a realized step and its 24
associated random steps formed a stratum of 25 steps that received a unique
identifier. Along all steps we extracted the above mentioned covariates (see
also Table S3). We also derived a binary covariate indicating whether or not a
step crossed a road and we identified the average distance of each step to the
nearest nearest road and the nearest source of water. We square-rooted the
values for \textit{distance to roads} and \textit{distance to water} to render a
decreasing marginal impact of distance. We standardized extracted values using a
z-score transformation, and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection model was then based on the assumption that dispersing wild dogs
assigned a selection score \(w(x)\) of the following exponential form to each
realized and random step \citep{Fortin.2005, Thurfjell.2014, Avgar.2016}:

@@

\subsection{Least-Cost Paths \& Least-Cost Corridors}
Our least-cost analysis revealed three major movement corridors of which all
were well-within the KAZA-TFCA boundaries (\Cref{LeastCost}). Furthermore, we
found little to no direct connectivity between peripheral points; that is, most
paths and corridors connecting two adjacent peripheral points run through more
central regions before heading towards their destination at the periphery
(\Cref{LeastCost}). One major corridor runs SE-NW and connects the
Okavango-Linyanti ecosystem in Botswana with Luengue-Luiana NP in Angola. A
second corridor runs W-E between Chobe NP in Botswana and Zimbabwe's Hwange NP.
A third major corridor runs NE-SW, completely across unprotected areas, and
connects Kafue NP in Zambia with more central regions of the KAZA-TFCA. Several
minor corridors branch off from these three major corridors; these include a
southward connection between the Okavango-Linyanti and the Central Kalahari Game
Reserve, a southwesterly corridor connecting Luengue-Luiana NP with Namibia's
Kaudom NP, and a northeasterly extension of the Hwange corridor into Zimbabwe's
Matusadona NP. According to our predictions, the corridor towards Kafue NP and
the corridor passing the Okavango-Linyanti region are the highest frequented
routes within the KAZA-TFCA (\Cref{LeastCost}, dark red colors). Our model did
not detect any significant direct corridors between Zimbabwe and Zambia or
Zambia and Angola, and only a very limited W-E direct connection between the
Okavango region and Namibia's Khaudum NP. Except for the corridor into the
Central Kalahari National Park, our model did not detect any significant
connectivity outside the boundaries of the KAZA-TFCA.

@@

\section{Discussion}
We used GPS relocation data collected on dispersing African wild dogs to
investigate whether their main movement corridors are contained within the
boundaries of the world's largest transboundary conservation area, namely the
KAZA-TFCA. Our analysis suggests that the KAZA-TFCA indeed encompasses all major
movement corridors of African wild dogs, demonstrating the potential value of
such an initiative. We thus expemplified how pertinent dispersal data of a
highly mobile species can be used to assess the adequacy of already existing or
planned protected areas. Our approach is neither limited to the African wild
dog, nor to our study area. All covariates used throughout this study are
readily available on a global scale and many of them are likely to be important
determinants of movement behavior, landscape permeability, and connectivity for
other species \citep{Thurfjell.2014, Zeller.2012}. Interestingly, our predicted
network of least cost-paths and corridors for African wild dogs shows surprising
similarities to corridors of dispersing lions inhabiting the same ecosystem
\citep{Elliot.2014, Cushman.2018}. This not only reinforces confidence in our
own predictions but also suggests potential synergies for the conservation of
these two, and possibly more, species. Expanding this analytical framework to
additional species will yield important insights on the consistency of
inter-specific movement corridors, thus highlighting areas that are
exceptionally valuable for the conservation of several species.

Locally, we identified the Okavango-Linyanti region as a potential dispersal hub
through which dispersersing wild dogs gain access to more peripheral regions of
the KAZA-TFCA. While this might partly be owed to the area's central position
within our study area, it is likely also due to its unique environmental and
anthropogenic characteristics. It appears that the absence of human activities
and the presence of relatively impermeable water bodies, such as the Okavango
Delta and the Linyanti Swamp funnel dispersal movements, resulting in a highly
frequented corridor. The region is also instrumental for connecting wild dog
subpopulations in Zambia's Kafue and Zimbabwe's Hwange NPs
\citep{Woodroffe.2012}. These subpopulations are currently separated by the
Zambezi River, forcing dispersers to detour via northern Botswana to reach more
peripheral areas. The key role of the Okavango-Linyanti region for overall
connectivity within the KATA-TFCA thus calls for actions to secure its
protection status in the future. While the region is currently a Wildlife
Management Area devoted to photographic tourism, it has neither the status of a
National Park nor that of a Game Reserve and so potentially faces growing
pressure towards exploitation of its natural resources under the influences of
an expanding human population, farming, and cropping industry. A similar case of
non-formally protected but key dispersal landscape is represented by the area
south of Kafue NP in Zambia, for which a disruption of its main and narrow
dispersal corridor would result in considerable isolation of its subpopulations.
We also revealed a potential southwards corridor between the Okavango-Linyanti
ecosystem and the Central Kalahari National Park. \cite{Elliot.2014} and
\citep{Cushman.2018} identified a similar corridor for dispersing lions,
suggesting that upholding and protecting a link between those ecosystems is
pivotal. Some areas through which the corridor runs are neither part of the
KAZA-TFCA nor profit from any form of protected status. In fact, human presence
and activities along the national road that longitudinally traverses this
corridor may limit realized connectivity \citep{Cozzi.2020}. Finally, our
connectivity map suggested that connectivity within the extent of the KAZA-TFCA
gradually decreases towards the borders of our study area, yet this might partly
be a result of boundary effects \citep{Koen.2010}.

@@

Our results further emphasize that human presence and activities constitute some
of the main barriers to connectivity. Our result that human influence is
detrimental to dispersal conforms to findings on dispersing wild dogs from
eastern Africa \citep{Masenga.2016, Oneill.2020} but conflicts with findings by
\cite{DaviesMostert.2012}, who reported a comparably high willingness of
dispersers to cross human-dominated landscapes in South Africa. We believe that
such differences are due to the unavailability of alternative ``natural''
routes, which may have forced dispersers in South Africa to cross human
dominated landscapes despite a strong aversion to do so. These contrasting
findings may hint at a considerable ability of dispersing wild dogs to adapt to
prevailing habitat conditions (see e.g. \cite{Woodroffe.2011}) and could imply a
limited model transferability between dispersersers from our study system to
dispersers found in other ecosystems, such as in South Africa
\citep{Whittington.2011, DaviesMostert.2012} or Kenya \citep{Woodroffe.2019,
Oneill.2020}. However, we believe that our study population adequately
represents wild dog populations residing within the greater extent of the
KAZA-TFCA, where more natural dispersal routes are readily available. Successful
conservation of this species is thus strictly contingent on policymakers' and
local authorities' willingness and ability to provide and conserve areas that
remain free from anthropogenic pressures. This is not only paramount in light of
providing connectivity and facilitating dispersal, but also in terms of reducing
human-caused mortality during dispersal. In fact, previous studies have shown
that human-caused mortality during dispersal, especially in proximity to
settlements and roads, represents a major threat to the dispersal ability of
this species \citep{Woodroffe.2019, Cozzi.2020}.

Besides anthropogenic pressures, we identified water as another major obstacle
to dispersal. This corroborates earlier studies which showed that water bodies
(including swamps) are almost impenetrable by resident packs
\citep{Abrahms.2017, Cozzi.2020} and only infrequently crossed by dispersing
individuals \citep{Cozzi.2020}. An accurate and dynamic representation of water
is thus imperative and particularly relevant in seasonal or flood-pulsing
ecosystems such as the Okavango Delta. In fact, future studies could investigate
how connectivity across such systems changes in result to seasonal changes in
the environment (see also \cite{Osipova.2019}).

Although dispersers avoided moving through water, we revealed a preference for
moving in proximity to water. This preference is likely explained by high
prey-availability in vicinity of water \citep{Western.1975, Bonyongo.2005}.
However, following the same reasoning, apex predators (e.g. lions, spotted
hyenas) and resident wild dogs may prefer proximity to water
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs into
prey-poorer areas that are more distant to water \citep{Ndaimani.2016} and pose
a lower risk of aggressive intraguild interactions \citep{Creel.1996,
Mills.1997, Webster.2012}. We could not control for the presence or absence of
apex predators or conspecifics during dispersal as we lacked corresponding data,
but these considerations may explain why we did not find a significant effect of
\textit{distance to water}. Given the influence that resident conspecifics,
predators, competitors, and prey can have on dispersers \citep{Cozzi.2018,
Armansin.2019} future studies should strive to collect and incorporate
corresponding ``social'' information into analyses of landscape connectivity to
more precisely reconstruct dispersal corridors.

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
\noindent 1. Conserving and managing large portions of land to connect wildlife
reserves is increasingly used to maintain and restore connectivity among
wildlife populations. Boundaries of such conservation areas are, however, often
determined based on expert opinion and socio-political constraints, but the
extent to which they match species' movement corridors is rarely examined. This
is mainly due to a lack of data, particularly on wide-ranging movement behavior
such as dispersal. Yet, empirically assessing the adequacy of protected areas is
key for the implementation of targeted management actions and use of limited
conservation funds.

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example of international conservation efforts. We used an integrated step
selection analysis to create a permeability surface for the KAZA-TFCA. We then
compared landscape permeability across different regions within the KAZA-TFCA as
well as outside its boundaries. Lastly, we calculated least-cost paths and
corridors to verify that major movement corridors are adequately encompassed
within the KAZA-TFCA.

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five countries. We further showed that major
movement corridors of wild dogs run within the KAZA-TFCA boundaries, although
some minor corridors remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly caused by different degrees
of human activities across regions, which hampered dispersal. Rivers and swamps
also limited dispersal, while other landscape features had a limited effect.

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing protected areas or could be used for
the planning of future such initiatives. Furthermore, the ability to reveal
regional differences in landscape permeability is fundamental for a coordinated
effort towards maintaining or restoring connectivity, especially where
international effort is required.
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become an utmost task for
conservation management \citep{Heller.2009, Doerr.2011, Rudnick.2012,
Cozzi.2020} and has resulted in an ever-growing number of large and often
transboundary protected areas \citep{Wolmer.2003}. Due to the very nature of
such large protected areas, a thorough quantitative assessment of their
effectiveness and adequacy of their boundaries is challenging
\citep{Rudnick.2012}. This challenge is complicated by a lack of animal movement
data at the appropriate spatial scale and an insufficient understanding of the
movement patterns and dispersal corridors of the species for which these
protected areas have been created \citep{Vasudev.2015, Tshipa.2017, Cozzi.2020}.
The use of subjective boundary assessments and the application of inappropriate
movement data, however, have revealed deficiencies in the past
\citep{Clevenger.2002, Pullinger.2010, Sawyer.2011, Elliot.2014, Abrahms.2016}.
Thus, an empirical assessment of the adequacy of already-existing or planned
protected areas using pertinent movement data of dispersing animals is paramount
for targeted use of valuable and scarce conservation funds
\citep{Pullinger.2010, Sawyer.2011}.

@@

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts, but for
which systematically collected GPS relocation data on dispersing individuals are
unavailable. Once widespread across the entire Sub-Saharan continent, wild dogs
have been widely extirpated through human persecution, habitat destruction, and
disease outbreaks \citep{Woodroffe.2012}. As a result, the species has become
one of Africa's most endangered large carnivores, and currently only survives in
small, spatially scattered subpopulations \citep{Woodroffe.2012}. Within these
subpopulations, wild dogs form cooperative breeding packs of up to thirty
individuals \citep{Frame.1979, Fuller.1992, Creel.2002}, whose social structure
is strongly governed by the process of dispersal \citep{McNutt.1996,
Woodroffe.2019, Behr.2020}. Both males and females disperse from their natal
pack, either alone or in same-sex dispersing coalitions, and search for
unrelated mates and a suitable territory to settle \citep{McNutt.1996,
Cozzi.2020, Behr.2020}. During dispersal, wild dogs can cover several hundred
kilometers, sometimes crossing national boundaries and areas that are strongly
influenced by humans \citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers during dispersal. The few studies that have
collected GPS relocation data during dispersal have shown that dispersers
quickly move over large distances \citep{Woodroffe.2019}, generally avoid
human-dominated landscapes \citep{Masenga.2016, Oneill.2020, Cozzi.2020} and
areas densely forested \citep{Oneill.2020}, but prefer proximity to water and
dirt roads \citep{Oneill.2020}.

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
used it to assess the adequacy of the KAZA-TFCA in securing connectivity for
this species. To achieve this, we estimated the degree of selection or avoidance
for environmental and anthropogenic landscape features. We then applied the
obtained habitat preferences and predicted a permeability surface spanning the
entire KAZA-TFCA and we investigated how landscape permeability varies
regionally and internationally and compared permeability within and outside the
KAZA-TFCA boundaries. Finally, we calculated least-cost paths and corridors to
identify major movement corridors, pinpoint dispersal hubs, and verify that the
KAZA-TFCA successfully covers these features.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E; elevation ca.
1'030 m; red dotted rectangle in \Cref{StudyArea}a) was outlined by a
rectangular bounding box stretching over 1.3 Mio km\textsuperscript{2} and
encompassing the entire KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA lies in the
basins of the Okavango and Zambezi rivers and includes parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia. With a total area of over 520'000
km\textsuperscript{2} it covers several already-existing national parks and
other protected areas. It constitutes the earth's largest transboundary
conservation area and is characterized by diverse landscapes, including savanna,
grassland, and dry or moist woodland habitats. Rainfall in the study area is
seasonal and lasts from November to March, whereas the rest of the year remains
dry \citep{Mendelsohn.2010}. The KAZA-TFCA also encloses the Okavango Delta,
which is one of the main strongholds of African wild dogs in southern Africa and
likely acts as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012}. The Okavango Delta represents a highly dynamic
hydrological flood-pulsing system \citep{McNutt.1996, Wolski.2017}. The extent
of the flood in the delta greatly changes within and between years depending on
the amount of rain that descends from the catchment areas in Angola and reaches
the distal ends of the delta between July and August (see Figure S5 for an
illustration of a typical flood pulse). The flood drastically affects
surrounding landscapes, so that during maximum extent (ca. 12'000
km\textsuperscript{2}) the delta becomes a patchy conglomerate of swamps, open
water, and islands, whereas these structures run dry when the flood retracts to
its minimum extent (ca. 5'000 km\textsuperscript{2}; \citealp{Gumbricht.2004,
Wolski.2017, McCarthy.1998}). Despite 36 NPs and other protected areas, there is
considerable human influence in some regions of the KAZA-TFCA, mainly
originating from farms, high human density, and road traffic.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' selection or avoidance for the above-mentioned spatial
covariates. In the iSSF framework, GPS relocations are coerced to  ``steps'',
where a step is defined as the connecting line between two consecutive GPS
relocations \citep{Turchin.1998}. Subsequently, habitat preferences are
estimated by contrasting covariates along realized steps to the same covariates
along alternative ``random''  steps \citep{Fortin.2005, Thurfjell.2014,
Avgar.2016}. Thus, we paired each realized step with 24 random steps. We
generated random steps by sampling turning angles from a uniform distribution
U(\(-\pi, +\pi\)) and step lengths from a gamma distribution that was fitted
using realized steps \citep{Avgar.2016}. Together, a realized step and its 24
associated random steps formed a stratum of 25 steps that received a unique
identifier. Along all steps we extracted the above mentioned covariates (see
also Table S3). We also derived a binary covariate indicating whether or not a
step crossed a road and we identified the average distance of each step to the
nearest nearest road and the nearest source of water. We square-rooted the
values for \textit{distance to roads} and \textit{distance to water} to render a
decreasing marginal impact of distance. We standardized extracted values using a
z-score transformation and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection model was then based on the assumption that dispersing wild dogs
assigned a selection score \(w(x)\) of the following exponential form to each
realized and random step \citep{Fortin.2005, Thurfjell.2014, Avgar.2016}:

@@

\subsection{Least-Cost Paths \& Least-Cost Corridors}
Our least-cost analysis revealed three major movement corridors of which all
were well-within the KAZA-TFCA boundaries (\Cref{LeastCost}). Furthermore, we
found little to no direct connectivity between peripheral points; that is, most
paths and corridors connecting two adjacent peripheral points run through more
central regions before heading towards their destination at the periphery
(\Cref{LeastCost}). One major corridor runs SE-NW and connects the
Okavango-Linyanti ecosystem in Botswana with Luengue-Luiana NP in Angola. A
second corridor runs W-E between Chobe NP in Botswana and Zimbabwe's Hwange NP.
A third major corridor runs NE-SW, completely across unprotected areas, and
connects Kafue NP in Zambia with more central regions of the KAZA-TFCA. Several
minor corridors branch off from these three major corridors; these include a
southward connection between the Okavango-Linyanti and the Central Kalahari Game
Reserve, a southwesterly corridor connecting Luengue-Luiana NP with Namibia's
Kaudom NP, and a northeasterly extension of the Hwange corridor into Zimbabwe's
Matusadona NP. According to our predictions, the landscapes in the
Okavango-Linyanti region are the highest frequented dispersal routes within the
KAZA-TFCA (\Cref{LeastCost}b). Our model did not detect any significant direct
corridors between Zimbabwe and Zambia or Zambia and Angola, and only a very
limited W-E direct connection between the Okavango region and Namibia's Khaudum
NP. Except for the corridor into the Central Kalahari National Park, our model
did not detect any significant connectivity outside the boundaries of the
KAZA-TFCA.

@@

\section{Discussion}
We used GPS relocation data collected on dispersing African wild dogs to
investigate whether their main movement corridors are contained within the
boundaries of the world's largest transboundary conservation area, namely the
KAZA-TFCA. Our analysis suggests that the KAZA-TFCA indeed encompasses all major
movement corridors of African wild dogs, demonstrating the potential value of
such an initiative. We thus expemplified how pertinent dispersal data of a
highly mobile species can be used to assess the adequacy of already existing or
planned protected areas. Our approach is neither limited to the African wild
dog, nor to our study area and thus applicable to any study system. All
covariates used throughout this study are readily available on a global scale
and many of them are likely to be important determinants of movement behavior,
landscape permeability, and connectivity for other species
\citep{Thurfjell.2014, Zeller.2012}. Interestingly, our predicted network of
least cost-paths and corridors for African wild dogs shows surprising
similarities to corridors of dispersing lions inhabiting the same ecosystem
\citep{Elliot.2014, Cushman.2018}. This not only reinforces confidence in our
own predictions but also suggests potential synergies for the conservation of
these two, and possibly more, species. Expanding this analytical framework to
additional species will yield important insights on the consistency of
inter-specific movement corridors, thus highlighting areas that are
exceptionally valuable for the conservation of several species.

Our results emphasize that human presence and activities constitute some of the
main barriers to connectivity. Our result that human influence is detrimental to
dispersal conforms to findings on dispersing wild dogs from eastern Africa
\citep{Masenga.2016, Oneill.2020} but conflicts with findings from South Africa
by \cite{DaviesMostert.2012}, who reported a comparably high willingness of
dispersers to cross human-dominated landscapes. We believe that such differences
are due to the unavailability of alternative routes through natural landscapes,
which may have forced dispersers in South Africa to cross human dominated
landscapes despite a strong aversion to do so. In this regard, our
representation of dispersal corridors and the resulting connectivity appear
conservative, as dispersers may be able to make the best out of a bad situation
and cross landscapes characterized by considerably unfavorable conditions
\citep{Palomares.2000, Elliot.2014}. Successful conservation of this species is
thus strictly contingent on policymakers' and local authorities' willingness and
ability to provide and conserve areas that remain free from anthropogenic
pressures. This is not only paramount in light of providing connectivity and
facilitating dispersal, but also in terms of reducing human-caused mortality
during dispersal. In fact, previous studies have shown that human-caused
mortality during dispersal, especially in proximity to settlements and roads,
represents a major threat to the dispersal ability of this species
\citep{Woodroffe.2019, Cozzi.2020}.

Besides anthropogenic pressures, we identified water as another major obstacle
to dispersal. This corroborates earlier studies which showed that water bodies
(including swamps) are almost impenetrable by resident packs
\citep{Abrahms.2017} and only infrequently crossed by dispersing individuals
\citep{Cozzi.2020}. An accurate and dynamic representation of water is thus
imperative and particularly relevant in seasonal or flood-pulsing ecosystems
such as the Okavango Delta. In fact, future studies could investigate how
connectivity across such systems changes in result to seasonal changes in the
environment (see also \cite{Osipova.2019}). Although dispersers avoided moving
through water, they selected locations in its vicinity. This preference may be
caused by the presence of prey close to water \citep{Western.1975,
Bonyongo.2005}. For the same reason, however, competitors such as lions, spotted
hyenas, and resident wild dogs may also use areas close to water sources
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs to move
into prey-poorer areas away from water \citep{Ndaimani.2016}. We could not
control for the presence or absence of apex predators or conspecifics during
dispersal as we lacked corresponding data, but these considerations may explain
why we did not find a significant effect of \textit{distance to water}. Given
the influence that resident conspecifics, predators, competitors, and prey can
have on dispersers \citep{Cozzi.2018, Armansin.2019} future studies should
strive to collect and incorporate corresponding ``social'' information into
analyses of landscape connectivity to more precisely reconstruct dispersal
corridors.

Locally, we identified the Okavango-Linyanti region as a potential dispersal hub
through which dispersing wild dogs gain access to more peripheral regions of the
KAZA-TFCA. While this might partly be owed to the area's central position within
our study area, it is likely also due to its unique environmental and
anthropogenic characteristics. It appears that the absence of human activities
and the presence of relatively impermeable water bodies, such as the Okavango
Delta and the Linyanti Swamp funnel dispersal movements, resulting in a highly
frequented corridor. The region is also instrumental for connecting wild dog
subpopulations in Zambia's Kafue and Zimbabwe's Hwange NPs
\citep{Woodroffe.2012}. These subpopulations are currently separated by the
Zambezi River, forcing dispersers to detour via northern Botswana to reach more
peripheral areas. The key role of the Okavango-Linyanti region for overall
connectivity within the KAZA-TFCA thus calls for actions to secure its
protection status in the future. While the region is currently a Wildlife
Management Area devoted to photographic tourism, it has neither the status of a
National Park nor that of a Game Reserve and so potentially faces growing
pressure towards exploitation of its natural resources under the influences of
an expanding human population, farming, and cropping industry. A similar case of
non-formally protected but key dispersal landscape is represented by the area
south of Kafue NP in Zambia, for which a disruption of its main and narrow
dispersal corridor would result in considerable isolation of its subpopulations.
We also revealed a potential southwards corridor between the Okavango-Linyanti
ecosystem and the Central Kalahari National Park. \cite{Elliot.2014} and
\citep{Cushman.2018} identified a similar corridor for dispersing lions,
suggesting that upholding and protecting a link between those ecosystems is
pivotal. Some areas through which the corridor runs are neither part of the
KAZA-TFCA nor profit from any form of protected status. In fact, human presence
and activities along the national road that longitudinally traverses this
corridor may limit realized connectivity \citep{Cozzi.2020}. Finally, our
connectivity map suggested that connectivity within the extent of the KAZA-TFCA
gradually decreases towards the borders of our study area, yet this might partly
be a result of boundary effects \citep{Koen.2010}.

@@

Our work shows how dispersal data of a highly mobile species can be used to
identify movement corridors and assess the adequacy of protected areas. In our
case, the predicted dispersal corridors of African wild dogs were well contained
within the boundaries of the world's largest transboundary conservation area,
namely the KAZA-TFCA, suggesting that it will significantly contribute to the
long-term viability of this species. Moreover, our connectivity network allowed
revealing potential dispersal hubs through which dispersers gain access to more
remote regions in the study area. Finally, our investigations showed that human
influence constitutes one of the main barriers to dispersal and substantially
reduces landscape connectivity accross large scales. Successful conservation of
wide-ranging species, such as exemplified by the African wild dog, will
therefore be contingent on the willingness of local authorities, policymakers,
and land managers to preserve areas that remain free from human strains.
Ultimately, our work contributes to the growing field of connectivity literature
and provides and easily expandable framework for the assessement of existing or
planned protected areas.
